% CONTENT GUIDANCE (delete before submission):
% - One paragraph, 150-250 words.
% - Lead with the problem (memory fragmentation in hybrid Mamba-Transformer
%   serving), the core idea (asymmetric virtual paging with dynamic
%   capacity rebalancing), and the headline result (cross-workload OOM
%   reduction vs the strongest static baseline, see
%   tables/generated/table_baseline_comparison.tex once generated).
% - Do not cite from the abstract; reserve citations for the body.

\begin{abstract}
Hybrid Mamba-Transformer language models combine attention with State Space Models (SSMs), creating asymmetric memory pressure between linearly scaling Key-Value (KV) blocks and fixed-size SSM states. Existing allocator designs for hybrid models typically either pad SSM state into KV-like pages or statically partition memory, which can waste capacity or fail to adapt under shifting workload distributions. Unified pools pad SSM states to match KV pages, causing severe capacity overestimation, while static dual-pool partitions cannot adapt to shifting prompt distributions. We present Asymmetric Virtual Memory Paging (AVMP), a memory allocator that decouples virtual address spaces across physically heterogeneous pools and dynamically rebalances capacity at runtime. We evaluate AVMP across a strictly deterministic 270-cell synthetic sweep plus a 60-cell ShareGPT-Vicuna trace replay, with paired-bootstrap 95\% confidence intervals on every headline claim. The dynamic AVMP allocator records goodput improvements from $1.83\times$ to $13.3\times$ on three synthetic workloads (CIs $[1.42, 2.60]$ to $[11.2, 16.0]$) and $2.36\times$ on ShareGPT (CI $[1.33, 5.65]$), plus a 7.6\% cross-workload Out-of-Memory (OOM) reduction (CI $[3.0\%, 12.7\%]$) against the best static baseline. A wall-clock phase decomposition shows the speedup splits into two mechanisms: faster OOM recovery on workloads with capacity pressure and faster per-call allocator service on workloads without it. Our logical operation counter guarantees byte-identical reproducibility for event-deterministic fields across reruns. We scope this design study as a pure Python prototype evaluated on a single GPU.
\end{abstract}
