\section{Related Work}
\label{sec:related}

\subsection{Cache Management for LLM Inference}
\label{sec:related:cache}

Production inference engines optimize transformer serving through paged memory allocation. Systems like PagedAttention and vLLM \cite{kwon2023pagedattention} manage the KV cache by dividing sequences into fixed-size blocks, reducing fragmentation and enabling efficient scatter-gather memory access. High-performance kernel libraries, such as FlashInfer \cite{ye2024flashinfer}, and production execution engines, like TensorRT-LLM, adopt similar block-based abstractions. These systems handle a single, homogeneous cache type well. They assume uniform memory access patterns and predictable block scaling tied directly to sequence length. AVMP extends this page-based paradigm to support two physically heterogeneous cache types. AVMP complements existing attention managers by adding a second pool type, not replacing them.

\subsection{Hybrid Architecture Serving}
\label{sec:related:hybrid}

Recent systems extend inference engines to support hybrid architectures containing both State Space Model and transformer layers \cite{dao2024mamba2, lieber2024jamba, glorioso2024zamba2, ren2024samba, dong2024hymba}. SGLang provides production serving for these models using a static dual-pool approach \cite{zheng2024sglang}. It provisions a \texttt{HybridReqToTokenPool} for attention and a \texttt{HybridLinearKVPool} for SSM state. Operators configure the partition at engine startup via the \texttt{mamba\_full\_memory\_ratio} parameter. This rigidity causes capacity failures when prompt distributions shift. As we demonstrate in \S\ref{sec:eval}, a default 0.9 ratio triggers 1221 OOM events compared to 552 OOMs at a 0.5 ratio on identical workloads.

Alternatively, vTensor introduces GPU virtual memory management for KV caches using hardware features to decouple physical memory mapping from virtual address spaces \cite{xu2024vtensor}. While vTensor targets a single cache type to reduce fragmentation, AVMP applies this virtual abstraction to two heterogeneous pools. AVMP contributes a unified virtual handle system combined with runtime capacity rebalancing, addressing both the correctness requirements and the adaptivity challenges in hybrid serving.

\subsection{Dynamic Resource Allocation}
\label{sec:related:dynamic}

Inference engines use runtime adaptation to maximize hardware utilization. Continuous batching strategies adapt batch sizes dynamically at iteration boundaries to increase throughput \cite{yu2022orca}. AVMP adapts pool capacity at request-level boundaries, using a more conservative trigger placed strictly in the allocation path. Memory eviction strategies, such as sliding window attention and attention sink retention \cite{xiao2024streamingllm}, discard older tokens dynamically under memory pressure. Token-importance heuristics like H2O retain heavy-hitter tokens \cite{zhang2023h2o}. These techniques operate within a single pool; AVMP migrates capacity across two heterogeneous pools.

At the framework level, tools like the PyTorch \texttt{expandable\_segments} allocator manage dynamic segment growth directly in CUDA. This allocator operates below the inference engine abstractions. AVMP operates at the pool level, translating application-level \texttt{CapacityError} exceptions into physical capacity migrations between distinct cache implementations. AVMP's CapacityError-triggered migration is conservative: it fires only when allocation would otherwise fail, not preemptively on transient pressure. Recent work on workload-aware request placement, such as Splitwise \cite{patel2024splitwise}, separates prompt and decode phases across hardware to exploit phase-specific resource demands. AVMP is orthogonal to phase splitting and could combine with such schedulers.

\subsection{Virtual Memory for ML Serving}
\label{sec:related:virtmem}

AVMP can be understood as applying the virtual/physical decoupling approach of vAttention and vTensor to a heterogeneous two-pool setting, with added runtime capacity migration between pools. Recent work explores GPU virtual memory primitives for ML workloads. vAttention~\cite{prabhu2024vattention} uses CUDA virtual memory APIs (\texttt{cuMemMap}) to dynamically map physical pages without contiguous tensor reservation, addressing fragmentation in KV cache management. vTensor~\cite{xu2024vtensor} extends this approach to flexible tensor management with hardware virtual memory features. Both target a single homogeneous cache type. AVMP differs by managing two heterogeneous pools (KV and SSM) under one virtual address space and providing dynamic capacity rebalancing across them. The current AVMP prototype implements virtual handle indirection in software; integrating \texttt{cuMemMap}-backed physical mapping is future work to address the 2$\times$ VRAM overhead reported in \S\ref{sec:eval:limitations}.
